manuscript_outline

### 2025-06-12 docs/manuscript/paper.Rmd
Edit Phylogenetic section

```{r tail-frog, fig.align = 'center', fig.cap="Custom hatching phenology model for coastal tailed frogs (Ascaphus truei).Panel A represents the model fit and raw data used to generate the effective value model and panel B are the effective values for daily temperatures between 6 and 20 °C. A dashed line with a 0.01 increase for every degree increase in included for reference."}
# Brown 1975, Comparitive Biochemistry and Physicology Part A: Physiology
# https://www.sciencedirect.com/science/article/pii/030096297590033X

@@

## Ecological

@@

```{r frog-preds, fig.height=6 , fig.width=8, fig.align='center', fig.cap="Modeled changes in developmental timing for coastal tailed frogs (Ascaphus truei) from 1991 to 2020. Panel A: Marginal mean trend across all sites showing a significant decline in days to hatching over time. Panel B: Site-specific (COMID) trends in predicted days to hatch, colored by mountain range. Estimates are based on the best fitting model, with stream temperatures from @siegel2023 and developmental timing modeled using hatchR. Shaded ribbons represent 95% confidence intervals."}

@@

## Phylogenetic

@@

```{r, fig.height= 8 , fig.width= 8, fig.align = 'center', fig.cap="Custom hatch timing models developed for four North American frog species. Models are parameterized from Moore 1939 and phenology is predicted using a randomly generated temperature regime with mean 16 °C (Panel A) and mean 24 °C (Panel B). Panel C shows the effective values for different mean daily temperatures for each species, which are effectively species-specific linearized developmental reaction norms. Note the genetic x environment interactions in Panel C as temperatures warm and shown by the rank order change in species in Panel A compared with Panel B."}

### all the Rana's below refer to taxonomy from the paper, they are now genus Lithobates

@@

rana_s_data <- tibble(temp = c(9, 21 ,26),
                      days = c(15,13,4))

rana_s_data <- tibble(temp = c(10,15.3,18.5,19.9, 23.7),
                      hours = c(305,128,87,79,52)) |>
  mutate(days = ceiling(hours/24))

rana_s_mod <- fit_model(temp = rana_s_data$temp,
                        days = rana_s_data$days,
                        species = "Rana_sylvatica",
                        development_type = "hatch")

@@

rana_pip_data <- tibble(temp = c(15.3,18.6,19.8, 26),
                        hours = c(168,116,99,57)) |>
  mutate(days = ceiling(hours/24))

rana_pip_mod <- fit_model(temp = rana_pip_data$temp,
                          days = rana_pip_data$days,
                          species = "Rana_pipiens",
                          development_type = "hatch")

# Lithobates clamitans ----------------------------------------------------------
# from Moore 1939

rana_c_data <- tibble(temp = c(15,19.8, 25.3,33.4),
                      hours = c(263,112,62,45)) |>
  mutate(days = ceiling(hours/24))

rana_c_mod <- fit_model(temp = rana_c_data$temp,
                        days = rana_c_data$days,
                        species = "Lithobates_clamitans",
                        development_type = "hatch")

@@

rana_p_data <- tibble(temp = c(15.5,18.6, 19.9, 25.7),
                      hours = c(190,126,106,63)) |>
  mutate(days = ceiling(hours/24))

rana_p_mod <- fit_model(temp = rana_p_data$temp,
                        days = rana_p_data$days,
                        species = "Rana_palustris",
                        development_type = "hatch")

# Lithobates example --------------------------------------------------------------------

### Figure A Code
# run for temp 16 C
set.seed(123)
temperature_data_cold <- data.frame(temperatures = sort(rnorm(30, mean = 16, sd =2.5)),
                               dates = seq(ymd("2000-04-01"), ymd("2000-04-30"), by = "days"))

froggy_mods <- c(rana_s_mod$expression$expression,
                 rana_p_mod$expression$expression,
                 rana_c_mod$expression$expression,
                 rana_pip_mod$expression$expression)

froggy_hatch_cold <- map(froggy_mods,
                    predict_phenology,
                    data = temperature_data_cold,
                    dates = dates,
                    temperature = temperatures,
                    spawn.date = "2000-04-01") |>
  map_dbl("days_to_develop") |> 
  tibble()

@@

froggy_hatch_cold <- froggy_hatch_cold |> 
    mutate(species = c("Wood Frog", "Pickerel Frog", 
                       "Green Frog","Northern Leopard Frog" )) |> 
  arrange(desc(days_to_develop))

@@

froggy_hatch_cold <- froggy_hatch_cold |>
  mutate(start = ymd("2000-04-01"),
         stop = start + days_to_develop,
         index = seq(20.5,17.5, by = -1))

p2.1.a <- ggplot() +
  geom_point(data = temperature_data_cold, aes(x = dates, y = temperatures)) +
  geom_line(data = temperature_data_cold, aes(x = dates, y = temperatures)) +
  geom_rect(data = froggy_hatch_cold, aes(xmin = start, xmax = stop, ymax =index-.15, ymin = index-.5, fill = species)) +
  geom_label(data = froggy_hatch_cold, aes(x = start + (stop - start) / 1.45, y = (index -0.325), label = days_to_develop)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("darkolivegreen4","deepskyblue4", "grey23",  "purple4"), name = "Species") +
  theme_classic() +
  theme(legend.position = c(0.75, 0.25))

# run for temp 24 C
set.seed(123)
temperature_data_warm <- data.frame(temperatures = sort(rnorm(30, mean = 24, sd =2.5)),
                                    dates = seq(ymd("2000-04-01"), ymd("2000-04-30"), by = "days"))

froggy_hatch_warm <- map(froggy_mods,
                         predict_phenology,
                         data = temperature_data_warm,
                         dates = dates,
                         temperature = temperatures,
                         spawn.date = "2000-04-01") |>
  map_dbl("days_to_develop") |> 
  tibble()

@@

froggy_hatch_warm <- froggy_hatch_warm |> 
  mutate(species = c("Wood Frog", "Pickerel Frog", 
                     "Green Frog","Northern Leopard Frog" )) |> 
  arrange(desc(days_to_develop))

@@

froggy_hatch_warm <- froggy_hatch_warm |>
  mutate(start = ymd("2000-04-01"),
         stop = start + days_to_develop,
         index = seq(28.5,25.5, by = -1))

p2.1.b <- ggplot() +
  geom_point(data = temperature_data_warm, aes(x = dates, y = temperatures)) +
  geom_line(data = temperature_data_warm, aes(x = dates, y = temperatures)) +
  geom_rect(data = froggy_hatch_warm, aes(xmin = start, xmax = stop, ymax =index-.15, ymin = index-.5, fill = species)) +
  geom_label(data = froggy_hatch_warm, aes(x = start + (stop - start) / 1.45, y = (index -0.325), label = days_to_develop)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("darkolivegreen4","deepskyblue4", "grey23",  "purple4"), name = "Species") +
  theme_classic() +
  theme(legend.position = "none")

@@

temps <- c(6:25)
species = c("Wood Frog", "Pickerel Frog", 
                       "Green Frog","Northern Leopard Frog" )
ef_vals <- NULL

for( m in 1:length(froggy_mods)){
  mod = froggy_mods[m]
  frog = species[m]

for (x in temps) {
  ef <- eval(parse(text = mod))
  temp_df <- data.frame(temperature = x, 
                        effective_value = ef, 
                        frog_species = frog)
  ef_vals <- rbind(ef_vals, temp_df)

@@

p2.2 <- ef_vals |> tibble() |> 
  ggplot(aes(x = temperature, y = effective_value, color = frog_species)) +
  geom_line() +
  geom_point() +
  geom_abline(intercept = 0, slope = .01, linetype = "dashed") +
  labs(x = "Daily Average Temperature (°C)", y = "Effective Value") +
  scale_color_manual(values = c("darkolivegreen4","deepskyblue4", "grey23",  "purple4"), name = "Species") +
  theme_classic() +
  theme(legend.position = "none")

(p2.1.a + p2.1.b) / p2.2 + plot_annotation(tag_levels = 'A')
```

>>>>

```{r tail-frog, fig.width=8, fig.height=3, fig.align='center', fig.cap="Custom hatching phenology model for coastal tailed frogs (Ascaphus truei).Panel A represents the model fit and raw data used to generate the effective value model and panel B are the effective values for daily temperatures between 6 and 20 °C. A dashed line with a 0.01 increase for every degree increase in included for reference."}
# Brown 1975, Comparitive Biochemistry and Physicology Part A: Physiology
# https://www.sciencedirect.com/science/article/pii/030096297590033X

@@

## Ecological Appication: Spatial and Temporal Trends in Development

@@

```{r frog-preds, fig.height=3 , fig.width=8, fig.align='center', fig.cap="Modeled changes in developmental timing for coastal tailed frogs (Ascaphus truei) from 1991 to 2020. Panel A: Marginal mean trend across all sites showing a significant decline in days to hatching over time. Panel B: Site-specific (COMID) trends in predicted days to hatch, colored by mountain range. Estimates are based on the best fitting model, with stream temperatures from @siegel2023 and developmental timing modeled using hatchR. Shaded ribbons represent 95% confidence intervals."}

@@

## Phylogenetic Application: Species Differences in Developmental Rates / Reaction Norms

@@

**BMM-Version**

Closely related species often diverge in their thermal tolerances and developmental strategies as part of niche differentiation. These differences can be visualized through reaction norms that capture genotype-by-environment (G×E) interactions, often derived from common garden or reciprocal transplant experiments. While such experiments are ideal for isolating genetic and plastic components of trait variation, they are logistically demanding and rarely available for many taxa.

**hatchR** offers a complementary approach by generating species-specific developmental models from published thermal rearing studies and applying them to shared temperature regimes. While this does not replace controlled experiments, it enables estimation of thermal reaction norms for development across species, even when the original experiments were conducted independently. By visualizing differences in effective value slopes and phenological outcomes under shared conditions, **hatchR** can highlight how developmental traits vary with phylogeny and ecological strategy.

To demonstrate this approach, we leveraged data from @moore1939 to parameterize developmental models for four widely distributed North American frog species in the genus *Lithobates*: wood frog (*L. sylvaticus*), pickerel frog (*L. palustris*), northern leopard frog (*L. pipiens*), and green frog (*L. clamitans*). These species co-occur in parts of eastern North America but differ in their breeding phenology, habitat associations, and thermal niches. For example, wood frogs are early breeders in cold, ephemeral wetlands, while green frogs breed later in more permanent, warmer waters.

We applied the custom parameterized models to two hypothetical spring temperature regimes with mean daily temperatures of 16 °C (cool) and 24 °C (warm). Under the cool regime, wood frogs hatched earliest while green frogs lagged behind by up to eight days—nearly double the developmental duration (Figure \@ref(fig:litho-pheno)A). However, under the warm regime, predicted hatch dates clustered tightly, with all species hatching within one day of each other (Figure \@ref(fig:litho-pheno)B). The effective value slopes (Figure \@ref(fig:litho-pheno)C) reveal these underlying differences in developmental response curves, serving as species-specific linearized reaction norms. Notably, the rank order of developmental timing reverses across the two thermal regimes, a hallmark of strong G×E interaction.

```{r litho-pheno, fig.height=8 , fig.width=8, fig.align='center', fig.cap="Custom hatch timing models developed for four North American *Lithobates* frog species. Models are parameterized from @moore1939 and phenology is predicted using two synthetic temperature regimes: Panel A shows predictions under a cooler regime (mean = 16 °C), while Panel B shows predictions under a warmer regime (mean = 24 °C). Panel C displays the effective value curves for each species—linearized approximations of thermal reaction norms. The crossover in rank order between Panels A and B illustrates genotype-by-environment interactions, reflecting species-specific thermal niches and breeding strategies."}
### all the Rana's below refer to taxonomy from the paper, they are now genus Lithobates

@@

rana_s_data <- tibble(
  temp = c(9, 21, 26),
  days = c(15, 13, 4)
)

rana_s_data <- tibble(
  temp = c(10, 15.3, 18.5, 19.9, 23.7),
  hours = c(305, 128, 87, 79, 52)
) |>
  mutate(days = ceiling(hours / 24))

rana_s_mod <- fit_model(
  temp = rana_s_data$temp,
  days = rana_s_data$days,
  species = "Rana_sylvatica",
  development_type = "hatch"
)

@@

rana_pip_data <- tibble(
  temp = c(15.3, 18.6, 19.8, 26),
  hours = c(168, 116, 99, 57)
) |>
  mutate(days = ceiling(hours / 24))

rana_pip_mod <- fit_model(
  temp = rana_pip_data$temp,
  days = rana_pip_data$days,
  species = "Rana_pipiens",
  development_type = "hatch"
)

# Lithobates clamitans --------------------------------------------------------
# from Moore 1939

rana_c_data <- tibble(
  temp = c(15, 19.8, 25.3, 33.4),
  hours = c(263, 112, 62, 45)
) |>
  mutate(days = ceiling(hours / 24))

rana_c_mod <- fit_model(
  temp = rana_c_data$temp,
  days = rana_c_data$days,
  species = "Lithobates_clamitans",
  development_type = "hatch"
)

@@

rana_p_data <- tibble(
  temp = c(15.5, 18.6, 19.9, 25.7),
  hours = c(190, 126, 106, 63)
) |>
  mutate(days = ceiling(hours / 24))

rana_p_mod <- fit_model(
  temp = rana_p_data$temp,
  days = rana_p_data$days,
  species = "Rana_palustris",
  development_type = "hatch"
)

# Lithobates example ----------------------------------------------------------

### Figure A Code
# run for temp 16 C
set.seed(123)
temperature_data_cold <- data.frame(
  temperatures = sort(rnorm(30, mean = 16, sd = 2.5)),
  dates = seq(ymd("2000-04-01"), ymd("2000-04-30"), by = "days")
)

froggy_mods <- c(
  rana_s_mod$expression$expression,
  rana_p_mod$expression$expression,
  rana_c_mod$expression$expression,
  rana_pip_mod$expression$expression
)

froggy_hatch_cold <- map(froggy_mods,
  predict_phenology,
  data = temperature_data_cold,
  dates = dates,
  temperature = temperatures,
  spawn.date = "2000-04-01"
) |>
  map_dbl("days_to_develop") |>
  tibble()

@@

froggy_hatch_cold <- froggy_hatch_cold |>
  mutate(species = c(
    "Wood Frog", "Pickerel Frog",
    "Green Frog", "Northern Leopard Frog"
  )) |>
  arrange(desc(days_to_develop))

@@

froggy_hatch_cold <- froggy_hatch_cold |>
  mutate(
    start = ymd("2000-04-01"),
    stop = start + days_to_develop,
    index = seq(20.5, 17.5, by = -1)
  )

p2.1.a <- ggplot() +
  geom_point(data = temperature_data_cold, aes(x = dates, y = temperatures)) +
  geom_line(data = temperature_data_cold, aes(x = dates, y = temperatures)) +
  geom_rect(data = froggy_hatch_cold, aes(xmin = start, xmax = stop, ymax = index - .15, ymin = index - .5, fill = species)) +
  geom_label(data = froggy_hatch_cold, aes(x = start + (stop - start) / 1.45, y = (index - 0.325), label = days_to_develop)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("darkolivegreen4", "deepskyblue4", "grey23", "purple4"), name = "Species") +
  theme_classic() +
  theme(legend.position = c(0.75, 0.25))

# run for temp 24 C
set.seed(123)
temperature_data_warm <- data.frame(
  temperatures = sort(rnorm(30, mean = 24, sd = 2.5)),
  dates = seq(ymd("2000-04-01"), ymd("2000-04-30"), by = "days")
)

froggy_hatch_warm <- map(froggy_mods,
  predict_phenology,
  data = temperature_data_warm,
  dates = dates,
  temperature = temperatures,
  spawn.date = "2000-04-01"
) |>
  map_dbl("days_to_develop") |>
  tibble()

@@

froggy_hatch_warm <- froggy_hatch_warm |>
  mutate(species = c(
    "Wood Frog", "Pickerel Frog",
    "Green Frog", "Northern Leopard Frog"
  )) |>
  arrange(desc(days_to_develop))

@@

froggy_hatch_warm <- froggy_hatch_warm |>
  mutate(
    start = ymd("2000-04-01"),
    stop = start + days_to_develop,
    index = seq(28.5, 25.5, by = -1)
  )

p2.1.b <- ggplot() +
  geom_point(data = temperature_data_warm, aes(x = dates, y = temperatures)) +
  geom_line(data = temperature_data_warm, aes(x = dates, y = temperatures)) +
  geom_rect(data = froggy_hatch_warm, aes(xmin = start, xmax = stop, ymax = index - .15, ymin = index - .5, fill = species)) +
  geom_label(data = froggy_hatch_warm, aes(x = start + (stop - start) / 1.45, y = (index - 0.325), label = days_to_develop)) +
  labs(x = "Date", y = "Temperature (°C)") +
  scale_fill_manual(values = c("darkolivegreen4", "deepskyblue4", "grey23", "purple4"), name = "Species") +
  theme_classic() +
  theme(legend.position = "none")

@@

temps <- c(6:25)
species <- c(
  "Wood Frog", "Pickerel Frog",
  "Green Frog", "Northern Leopard Frog"
)
ef_vals <- NULL

for (m in 1:length(froggy_mods)) {
  mod <- froggy_mods[m]
  frog <- species[m]

  for (x in temps) {
    ef <- eval(parse(text = mod))
    temp_df <- data.frame(
      temperature = x,
      effective_value = ef,
      frog_species = frog
    )
    ef_vals <- rbind(ef_vals, temp_df)
  }
}

@@

p2.2 <- ef_vals |>
  tibble() |>
  ggplot(aes(x = temperature, y = effective_value, color = frog_species)) +
  geom_line() +
  geom_point() +
  geom_abline(intercept = 0, slope = .01, linetype = "dashed") +
  labs(x = "Daily Average Temperature (°C)", y = "Effective Value") +
  scale_color_manual(values = c("darkolivegreen4", "deepskyblue4", "grey23", "purple4"), name = "Species") +
  theme_classic() +
  theme(legend.position = "none")

(p2.1.a + p2.1.b) / p2.2 + plot_annotation(tag_levels = "A")
```

This example demonstrates how **hatchR** can be used to explore evolutionary and ecological differences in thermal sensitivity across species. When combined with phylogenetic or trait-based analyses, such reaction norms can reveal how developmental strategies align with species’ ecological niches, helping to explain the persistence of closely related species across thermal gradients.
